Typed Functional Genetic Programming

### 2013-07-31 tex/thesis.tex
čtvrtek 0:23

% možná místo classical GP řikat standard GP jako montana95

@@

\newcommand{\EA}{EA\xspace} % TODO nahradit správnym "evoluťionar algorithms"

@@

Supervisor of the master thesis: & RNDr. Petr Pudlák, Ph.D. \\
\noalign{\vspace{2mm}}
Study programme: & Theoretical Computer Science \\ %Teoretická informatika \\
\noalign{\vspace{2mm}}
Specialization: & 
%Neprocedurální programování a umělá inteligence \\
{\fontsize{0.3cm}{1em}\selectfont 
%Neprocedurální programování a umělá inteligence} \\
Non-Procedural Programming and Artificial Intelligence} \\
\end{tabular}

@@

\noindent
$<$dedication$>$

@@

Název práce:
Název práce
% přesně dle zadání

Autor:
Jméno a příjmení autora

Katedra:  % Případně Ústav:
Název katedry či ústavu, kde byla práce oficiálně zadána
% dle Organizační struktury MFF UK

Vedoucí diplomové práce:
Jméno a příjmení s tituly, pracoviště
% dle Organizační struktury MFF UK, případně plný název pracoviště mimo MFF UK

@@

Klíčová slova:
% 3 až 5 klíčových slov

@@

Title:
% přesný překlad názvu práce v angličtině

Author:
Jméno a příjmení autora

Department:
Název katedry či ústavu, kde byla práce oficiálně zadána
% dle Organizační struktury MFF UK v angličtině

Supervisor:
Jméno a příjmení s tituly, pracoviště
% dle Organizační struktury MFF UK, případně plný název pracoviště
% mimo MFF UK v angličtině

@@

Keywords:
% 3 až 5 klíčových slov v angličtině

@@

GP constructs the population of solution programs from 
two sets of building blocks; set $F$ of functions 
and set $T$ of constants and variables. 
In classical GP \textit{closure} of those 
building blocks is required.
This means that all functions in $F$ must accept all 
constants and variables from $T$ as their arguments and produce 
such values that are also acceptable by all functions in $F$ as arguments.
Main motivation behind this thesis is to design and implement a system that
eliminates this constraint. 
This can be achieved by utilization of \textit{types}.\\

Utilization of types in GP is not a new idea; 
it was dealt with in various previous works 
\cite{yu01,montana95,haynes96,olsson94}.
The contribution of this thesis can be seen 
in that it deals with this task using concepts from
theory of typed $\lambda$-calculus and combinatory logic
and shows how to use them in GP.
Hopefully, it is also shown that those two fields are
nicely linked together, e.g., the concept of \textit{building blocks}
corresponds to concept of \textit{context}. \\

Notable highlights are: 
\begin{itemize}
 \item \textit{Inhabitation tree} (a slight modification 
   of \textit{inhabitation machine} presented in \cite{barendregt10}
   suited better for reasoning about implementation details) 
   is presented and used for \textit{generation of individuals}.
   This generation process is parameterized by simple 
   \textit{search strategy}. Several search strategies are presented,
   such as \textit{systematic} strategy and strategy performing
   classical GP individual generation.     
 \item \textit{Abstraction elimination} is utilized
   to enable use of simple \textit{mutation operator}.
% \item \textit{Polymorphic} generalization of inhabitation tree,
%   enabling use of polymorphic functions as building blocks, 
%   is presented.

@@

Design of a GP system utilizing these methods is presented.
This design aims to be generalization of classical GP 
developed by Koza \cite{koza92}, rather than being 
specialized variant of GP using more recent techniques.
This decision is based on belief
that simplicity of design of classical GP plays nicely
with testing those $\lambda$-calculus concepts in 
the evolutionary context and on belief that modern specialized 
techniques can be incorporated later.

%The implementation of our system aims 
Described system is implemented in purely functional language 
\textit{Haskell}. Evaluation of individual \lterms is 
performed by translating them to
Haskell programs and evaluating them in Haskell interpreter. 
The core part of the system runs as server, which is 
managed through user interface accessible via
web browser. Solution individuals are also translated
to \textit{JavaScript} expressions, which enables  
an interactive solution presentation.
This approach helps us with demonstrating 
that \lterms can be easily translated to a 
widely used programming language.

Several example problems for the designed system are presented 
and examined. Those examples are not intended to 
bring breakthrough results; they are intended to 
serve as demonstration of that the system is working properly
and as demonstration of several directions in which the system
can be used.

\begin{todo}
 & možná přidat ještě něco z poznámek co jsou teď v Conclusion
 & vlastně i koza hrotí evolution of constrained syntactic structures !!!
 & my tu řikáme že to závadíme kuli typum, ale my děláme v víc! -dáváme tam
   i abstrakce atd - ale mužeme říct, že tim že se to pak převede zase na 
   to bez abstrakcí tak setrváváme dal v krasnym bez-lambda abstrakčí ale
   s tou výhodou že naše metoda gerneruje všechny struktury pro ten danej typ ...
   nebo tak něco - asi to dát spíš do design of our system nebo někam
   ale někde to asi zmínit
\end{todo}

@@

Another crucial input besides fitness function is a collection of \textit{building blocks}.
It is collection of symbols (accompanied with an information about number of arguments).
Those symbols are used to construct trees representing individuals.\\

\Lets describe GP algorithm briefly:

At the beginning initial population is generated from building blocks randomly.

A step of GP algorithm is stochastic transformation of the current population into 	
the next population.

This step consists of two sub steps:
\begin{itemize} 
	\item Selection of \textit{parents} for individuals of the next population based on the fitness.
	      The bigger fitness of an individual of the current population is, 
	      the better chance of success being selected as parent it has.  
	\item Application of genetic operators (such as \textit{crossover}, 
	      \textit{reproduction} and \textit{mutation}) 
		  on parent individuals producing new individuals of the next population.  
\end{itemize}	  
This transformation is repeatedly applied for a predefined number of steps (which is called 
number of \textit{generations}) or until some predefined criterion is met.	
\\\\
\Lets now look on GP at more detail.

\section{Program trees}
\label{GP-prog-trees}

@@

\begin{samepage}
\begin{itemize}
	\item Constant or variable symbol $s$ is S-expression.\footnote{
	      By constants we also mean procedures with zero arguments.}
	\item Let there be a function with symbol $f$ which has $n$ arguments.

@@

	      Then ( $f$ $e_{1}$ ... $e_{n}$ ) is expression.\footnote{
	      This notation comes from Lisp programming language, 
	      a more standard notation would be $f(e_{1}, ... ,e_{n})$.}   
\end{itemize}
\end{samepage}

There is straightforward tree representation corresponding to these two cases:

\begin{itemize}
	\item One node tree $s$.
    \item \Tree
			[.$f$	
		 		\text{$t_{1}$}
		 		\text{...}
		 		\text{$t_{n}$} ]\\\\
		 Where $t_{1}, ..., t_{n}$ are trees corresponding 
		 to S-expressions $e_{1}, ..., e_{n}$.	   
\end{itemize}

@@

\begin{todo}
 & nějak to tu přepsat aby to bylo ve stejnym stylu jako v \ref{tree-reps}
\end{todo}

@@

We can say that the motivation behind this thesis is to construct system,
where we eliminate this constraint.

@@

We will describe tree generating method described by Koza 
called \textit{ramped half-and-half}  \cite{koza92}. \\

@@

\begin{todo}
 & Dozjistit a dopsat jak to Koza má s vyžadovánim unikátnosti při generování.
\end{todo}

@@

In the field of evolutionary algorithms there is plenty of 
options for selection mechanisms 
for us to choose from. Again we will describe mechanism Koza
used in his first book on GP. It is the \textit{Roulette selection}.
It uses fitness value of each individual in straightforward way to determine probability of selecting this individual.\\

Let there be $popSize$ individuals in the population.\\
And let $f_{i}$ be fitness value for individual i 
where $i \in \setDots{1}{popSize}$.

@@

In GP the most common crossover is \textit{Subtree swapping crossover}.
This crossover randomly selects one node in each parent tree.
Two new child individuals are constructed by swapping subtrees 
that have roots in those selected nodes.\\

@@

And here are two child trees with swapped subtrees:

@@

\begin{todo}
 & POZOR POZOR !!!!! zapoměl jsem tam že je 90\% procent šance 
   že se vybere internal node a jen 10\% šance že terminal node
\end{todo}

@@

\section{Construction of a next population}

@@

\begin{todo}
 & !!!
 & zmínit tu terminate bool
\end{todo}

@@

Bellow is described GP algorithm in pseudocode.

\Pseudokod{GP( fitness, \TuF, popSize, numGens, probabs )}{
  $gen \leftarrow$ 0 \;\;

  $pop \leftarrow generateInitialPopuletion( $ \TuF$ ) $ \;
  ($popWithF,terminate,best) \leftarrow$ 
  evaluate($fitness$, \TuF, $pop$)\;\;

@@

			\Case{Crossover}{
				$parent1 \leftarrow$ selection( $popWithF$ ) \;
				$parent2 \leftarrow$ selection( $popWithF$ ) \;\;

@@

				$newPop$.insert( $child1$ ) \;
				$newPop$.insert( $child2$ ) \;
				$i \leftarrow i + 2$ \;
			}
			\Case{Reproduction}{
				$indiv \leftarrow$ selection( $popWithF$ ) \;
				$newPop$.insert( $indiv$ ) \;
				$i \leftarrow i + 1$ \;
			}
			\Case{Mutation}{
				$indiv \leftarrow$ selection( $popWithF$ ) \;
				$mutant \leftarrow$ mutate( $indiv$ , \TuF ) \;
				$newPop$.insert( $mutant$ ) \;
				$i \leftarrow i + 1$ \;	
			}		
		}
	}\;

	$pop \leftarrow newPop$  \;
	($popWithF,terminate,best) \leftarrow$ 
	evaluate($fitness$, $TuF$, $pop$)\;
	$gen \leftarrow gen + 1$ \; 
 }\;

 \Return pop \;
}

@@

Behaviors of contained procedures
generateInitialPopuletion(),
evaluate(),
probabilisticallySelectOperation(),
selection(),
crossover() 
and mutation() are hopefully clear from verbal description.

\begin{todo}
 & mention best preservation and same popSize checking.
 & popsat co je to popWithF
 & popsat co je to terminate 
 & ten algoritmus v kozovi dělá "Designate Result" já tam vracim 
   poslední populaci
\end{todo}

@@

@@

\begin{todo}
 & Až to bude dostatečně sepsáno, tak do této kapitoly taky přidat
   část o abstraction elimination a o polymorfním zobecnění inhabitation tree.
   Nezapomenout to pak zmínit i v tomhle úvodu.
\end{todo}

@@

We use bold upper case symbols such as 
$\Scomb, \Kcomb, \Icomb$ as abbreviation
for specific terms, e.i. 
$\Icomb = \lamb{x}{x}$.  
\\[1em]

@@

$M[x:=N]$ the substitution of a term $N$ for the free occurrences of 
a variable $x$ in a term $M$ is defined inductively as follows.
\begin{align*}
x[x:=N]           &= N \\
y[x:=N]           &= y \text{ ~~~~where } x \not= y  \\
(P~Q)[x:=N]       &= ( P[x:=N]  ~ Q[x:=N] ) \\
\lamb{x}{P}[x:=N] &= \lam{x}{P}\\
\lamb{y}{P}[x:=N] &= \lam{y}{( P[x:=N] )} \text{ ~~~~where } x \not= y
\end{align*}

\begin{todo}
& možná zmínit že y nesmí být v N volně v připadě\\
  $\lamb{y}{P}[x:=N] = \lam{y}{( P[x:=N] )}$ , 
  ale řekl bych že zbytečnej detail
\end{todo}

@@

In other words, $\beta$-reduction is the process 
of insertion of arguments supplied to a function into 
its body.

\begin{todo}
& asi to přepsat nějak obecně pro redukci R a pak zadat beta,eta,eta-1 
  jako specialni připady
& !!!!!!! zmínit $=_\beta$ a $\twoheadrightarrow_\beta$  a i pro ostatní redukce!!!!!
& definovat \textbf{CR} a říct o tom lemátku, že má CR redukce má unikátní nf.
&& V \cite{barendregt84} (ta nová kniha s žlutejma deskama) je ukázano že je 
	 eta CR (Church Rosserovská)
& zmínit Theorem 2B.4 z \cite{barendregt10}
& co to znamená mít nf
\end{todo}

@@

$$\lamb{x}{(M~x)} \eArrow M \text{ ~~~~where } x \not\in FV(M) $$

@@

$$M \eeArrow \lamb{x}{(M~x)} \text{ ~~~~where } x \not\in FV(M) $$

%\subsection{Reductions for $c \in C$}
%\textbf{TODO možná vynechat...}

@@

A \lterm is a \textit{$\beta$-normal form} if it does not have a $\beta$-redex as
subterm.
\\\\
A normal form may be thought of as a result of a term evaluation.

@@

\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}
\newcommand{\atTree}{@-tree\xspace}

@@

\begin{todo}
 & napsat transformace mezi \atTree a \sexprTree
%& ujistit se že je v části o tree reprezentacích 
%  je napsaný co to znamená že podstrom má nějakej typ,
%  že podstromy odpovídaj podvýrazum atd
% & EXAMPLES of tree representations of \lterms  
% & nějak to sjednotit s tim uplně na začátku v popisu GP kde se popisuje
%   přímo \sexprTree a říct proč se tomu tak řiká 
%   (na tom místě co je v textu dřív) 
% & zkontrolovat že reference tree-reps se odkazuje na dobrý místo,
%   mohlo by se to rozbít tim že tree reprezentations - typovany/netypovany 
%   jsou od sebe oddelený
\end{todo}

@@

%\begin{definition}
%Let $\Gamma \in \mathfrak P \left({\Lambda \times  \mathbb{T}}\right)$. 
%($\Gamma$ is a set of statements of a form $M : \sigma$.)	
%Then $\Gamma$ is {\it context} if it obeys following 
%conditions.
%%\footnote{The $\pi_1$ corresponds to the projection of the 
%%first component of the Cartesian product.}
%\begin{align*}
%\forall (x,\sigma) \in \Gamma &: x \in V \cup C  
%&\textit{(x is variable or constant)}\\
%\forall (x_1,\sigma_1),(x_2,\sigma_2) \in \Gamma &: 
%x_1,x_2 \in V , x_1 \neq x_2 \then \sigma_1 \neq \sigma_2
%&\textit{(variables are distinct)}
%\end{align*}~
%Ten druhej je navíc blbě, má tam bejt celej stejtment se nerovná pak...
%\end{definition}

@@

In our system, the set of \textit{building blocks} 
(the equivalent of the set $T \cup F$ in classical GP)
will be a context with all declarations having \textit{constants}
as subjects, i.e., a set of constants together with their types.

In the process of term generation, 
this initial context is being further extend with variables
 --- by "entering inner scopes of bodies of lambda functions".

%\subsection{Statements of the form \GMS}

\begin{todo}
 & zapsat sem notaci s $\Gamma,x:\sigma $ denotes $ \Gamma\cup\{(x:\sigma)\}$ 
\end{todo}

%\subsection{Inference rules}
\subsection{Simply typed $\lambda$-calculus}

@@

\begin{todo}
   & říct že se vždycky zastavěj (strong normalization)
     && a jaký to má výhody pro nás v GP (duležitý),
     && potřebný definice pro to uvést v \lterms sekci
     && musí se říct co je strongly normalizing, 
        tzn simpli typed $lambda$-kalkulus termy a $\beta$-redukce /
        $\beta\eta$-redukce
%     && řekl bych že eta redukce je SN vždycky (tzn i untyped lk)
%        pač plati (CR+terminující pak SN) a je CR vzhledem k barendrecht-stará-žlutá
%        a terminující je rpotože zmenšuje term s nemuže zmenšovat donekonečna    
   & možná zmínit že se jedná o \textit{gentzen style proof}
   & říct že jsou to modus ponens a jeden směr věty o dedukci  
\end{todo}

@@

\begin{todo}
 & možná konec týhle subsekce trochu přeformulovat, je to takový kostrbatý...
\end{todo}

\section{Term generating grammar}

@@

For every non-terminal $(\sigma \ar \tau,\Gamma)$
and for every variable $x$
there is following grammar rule.\footnote{
\textbf{TODO !!!!} nějak okomentovat že by měla být freš,
pokudmožno v celym termu ale to že byse dělalo blě aby to
zůstala bezkontextová gramatika, ale je to celkem nepodtatnej detail
protože tyhle term generating gramatiky sou tu spíš
pro zajímavost a pro pevnější kramfleky, než pro praktický 
generování}

@@

$$ 
	( \sigma \rightarrow \tau , \Gamma )  \gar
	\bigg( \mbox{ {\Large $\lambda$ x . }}( \tau , \Gamma \cup \{ (x,\sigma) \} ) \quad \bigg)
$$\\

@@

\begin{todo}
   & buď dokázat ty převody na grammar rules explicitně
     nebo říct že se to dokáže podobnym stylem jako v dukazu 
     v sekci \ref{barlike}
   & možná ocitovat toho van Wijngaarden [1981] s těma 2-level grammar
     jak je to v Barendregtovi \cite{barendregt10} 
   & vyřešit jak to tu popsat s tou freš proměnnou x v introdukci...
     && stejně tak v tom obecnym případě to dělá neplechu
\end{todo}

@@

\begin{todo}
 & potřebuju eště nějaký \textbf{propositions} z \cite{barendregt10} tu uvíst ?
 & ta definice je tam daná pro $\tur{}{M}{\sigma}$,
   ukazat ze ta naše je ok, že tam má navíc to $\Gamma$.
   Stačí udělat tríček že beru funkci co má lambda hlavu odpovídající 
   tomu kontextu.
\end{todo}

@@

\begin{todo}
 & možná zmínit že podobně jako mužeme transformovat inference rules
   na grammar rules, mužeme i obráceně grammar rules
   převést na inference rules - to bylo ostatně uděláno v tom dukazu ale
   je to podle mě hezký říct ty pravidla explicitně.
\end{todo}

@@

\begin{todo}
& možná eště zmínit něco jako : \\
Jedna drobná nevýhoda je, že pokud chceme generovat termy od nejmenšího do největšího
a pokud lnf generujeme od nejmenšího po největší tak pak generovaní \benf
timhle trikem nemusí bejt od nejmenšího po nějvětší. V naší aplikaci (tzn GP)
ale nijak nebazírtujeme na týhle vlastnosti a nakonec budeme spíš stejně používat
nějaký nedeterminizmy aby to fungovalo líp, takže to vubec nevadí, že budou trochu přeházený. 
\end{todo}

@@

\Lets consider following definition of 
\textit{and-or tree}\footnote{
\textbf{TODO}: Mention that on [WIKI] there is more general definition, but 
for our purposes is this one sufficient.}.

@@

\newpage
\section{ Generating terms in \lnf by solving inhabitation tree }

@@

\begin{todo}
& skontrolovat ty dukazy pořádně sepsal sem 
  je dost rychle a mužou bejt divný. 
& A možná to ještě to nějak okomentovat
\end{todo}

>>>>

@@

\newcommand{\atTree}{@-tree\xspace}

%\newcommand{\EA}{EA\xspace} % TODO nahradit správnym "evoluťionar algorithms"

@@

Supervisor of the master thesis: & RNDr. Petr Pudlák, Ph.D. \\
\noalign{\vspace{2mm}}
Study programme: & Computer Science \\ %Teoretická informatika \\
\noalign{\vspace{2mm}}
Specialization: & Theoretical Computer Science
%Neprocedurální programování a umělá inteligence \\
%{\fontsize{0.36cm}{1em}\selectfont 
%Neprocedurální programování a umělá inteligence} \\
%Non-procedural Programming and Artificial Intelligence} \\
%}\\
\end{tabular}

@@

\noindent
%$<$dedication$>$
I would like to express my gratitude to my supervisor RNDr. Petr Pudlák, Ph.D.
for the useful comments, remarks and help with this master thesis.
I would also like to thank my parents, who have supported me throughout
my studies.

@@

Název práce:
Typed Functional Genetic Programming
% přesně dle zadání

Autor:
Tomáš Křen

Katedra:  % Případně Ústav:
Katedra teoretické informatiky a matematické logiky
% dle Organizační struktury MFF UK

Vedoucí diplomové práce:
RNDr. Petr Pudlák, Ph.D., 
Katedra teoretické informatiky a matematické logiky
%Jméno a příjmení s tituly, pracoviště
% dle Organizační struktury MFF UK, případně plný název pracoviště mimo MFF UK

@@

Klíčová slova:
geneticke programovani, funkcionalni programovani
% 3 až 5 klíčových slov

@@

Title:
Typed Functional Genetic Programming
% přesný překlad názvu práce v angličtině

Author:
Tomáš Křen

Department:
Department of Theoretical Computer Science and Mathematical Logic
% dle Organizační struktury MFF UK v angličtině

Supervisor:
RNDr. Petr Pudlák, Ph.D., 
Department of Theoretical Computer Science and Mathematical Logic
%Jméno a příjmení s tituly, pracoviště
% dle Organizační struktury MFF UK, případně plný název pracoviště
% mimo MFF UK v angličtině

@@

Keywords:
genetic programming, functional programming
% 3 až 5 klíčových slov v angličtině

@@

GP constructs the population of solution programs from 
two sets of building blocks; set $F$ of functions 
and set $T$ of constants and variables. 
In standard GP \textit{closure} of those 
building blocks is required.
This means that all functions in $F$ must accept all 
constants and variables from $T$ as their arguments and produce 
such values that are also acceptable by all functions in $F$ as arguments.
Main motivation behind this thesis is to design and implement a system that
eliminates this constraint. 
This can be achieved by utilization of \textit{types}.
Utilization of types in GP is not a new idea; 
it was dealt with in various previous works 
\cite{yu01,montana95,haynes96,olsson94}.\\

The contribution of this thesis can be seen 
in that it deals with this task using concepts from
theory of typed $\lambda$-calculus and combinatory logic
and shows how to use them in GP. 
Hopefully, it is also shown that those two fields are
nicely linked together.
%, e.g., the concept of \textit{building blocks}
%corresponds to concept of \textit{context}.
Several enhancements are derived from this theoretical background
and shown to have positive effect on performance of our system.  
Our main focus lies in the process of generating the initial population.

Design of a GP system utilizing these methods is presented.
This design aims to be generalization of standard GP 
developed by Koza\cite{koza92}, rather than being 
specialized variant of GP using more recent techniques.
This decision is based on belief
that simplicity of design of standard GP plays nicely
with testing those $\lambda$-calculus concepts in 
the evolutionary context and on belief that modern specialized 
techniques can be incorporated later.\\

Notable highlights of the thesis are: 
\begin{itemize}
 \item \textit{Inhabitation tree} (a slight modification 
   of \textit{inhabitation machine}\cite{barendregt10}
   suited better for reasoning about implementation details) 
   is presented and used for \textit{generation of individuals}.
   This generation process is parameterized by simple 
   \textit{search strategy}. Several search strategies are presented,
   such as \textit{systematic} strategy performing exhaustive enumeration
   of individuals and strategy performing same individual generation 
   as standard GP method \textit{ramped half-and-half}.  
 \item One of those search strategies, the \textit{geometric} strategy, is
   examined in a couple of experiments and shown to have 
   several desirable properties.
 \item The \textit{$\eta$-normalization} of generated individuals
       and \textit{\atTree} representation of individuals, 
       two enhancements of the system inspired from the theoretical
       background of the system are examined and shown to have 
       improving effect on performance of the system.     
 \item \textit{Abstraction elimination} is utilized
   to enable use of simple \textit{mutation operator}.
% \item \textit{Polymorphic} generalization of inhabitation tree,
%   enabling use of polymorphic functions as building blocks, 
%   is presented.

@@

%The implementation of our system aims 
Described system is implemented in purely functional language 
\textit{Haskell}. Evaluation of individual \lterms is 
performed by translating them to
Haskell programs and evaluating them in Haskell interpreter. 
The core part of the system runs as server, which is 
managed through user interface accessible via
web browser. Solution individuals are also translated
to \textit{JavaScript} expressions, which enables  
an interactive solution presentation.
This approach helps us with demonstrating 
that \lterms can be easily translated to a 
widely used programming language.\\

Several example problems for the designed system are presented 
and examined. Those examples are not intended to 
bring breakthrough results; they are intended to 
serve as demonstration of that the system is working properly,
as instrument for examining presented techniques and 
as demonstration of several directions in which the system
can be used.

@@

Another crucial input besides fitness function is a collection of 
\textit{building blocks}\footnote{Throughout this thesis, 
we use the term \textit{building blocks} 
differently than it is sometimes used in GP context. 
Hopefully it will not confuse the reader.}.
It is a collection of symbols (accompanied with an information about number of arguments).
Those symbols are used to construct trees representing individuals.\\

\Lets describe GP algorithm briefly.
At the beginning, initial population is generated from building blocks randomly.
A step of GP algorithm is stochastic transformation of the current population into 	
the next population.

This step consists of two sub steps:
\begin{itemize} 
	\item Selection of \textit{parents} for individuals of the next population based on the fitness.
	      The bigger fitness of an individual of the current population is, 
	      the better chance of success being selected as parent it has.  
	\item Application of genetic operators (such as \textit{crossover}, 
	      \textit{reproduction} and \textit{mutation}) 
		  on parent individuals producing new individuals of the next population.  
\end{itemize}	  
This transformation is repeatedly applied for a predefined number of steps (which is called 
number of \textit{generations}) or until some predefined criterion is met.	
\\\\
\Lets now look on GP in greater detail.

\newpage
\section{Program trees}
\label{GP-prog-trees}

@@

\begin{samepage}
\begin{enumerate}
	\item Constant or variable symbol $s$ is S-expression.\footnote{
	      By constants we also mean procedures with zero arguments.}
	\item Let there be a function with symbol $f$ which has $n$ arguments.

@@

	      Then ( $f$ $e_{1}$ ... $e_{n}$ ) is expression.\footnote{
	      This notation comes from Lisp programming language, 
	      a more standard notation would be $f(e_{1}, ... ,e_{n})$.}   
\end{enumerate}
\end{samepage}~\\
There is straightforward tree representation corresponding to these two cases:

\begin{enumerate}
	\item One node tree $s$.
    \item \mbox{\Tree
			[.$f$	
		 		\text{$t_{1}$}
		 		\text{...}
		 		\text{$t_{n}$} ]}\\\\
		 where $t_{1}, ..., t_{n}$ are trees corresponding 
		 to S-expressions $e_{1}, ..., e_{n}$.	   
\end{enumerate}

@@

@@

We can say that the motivation behind this thesis is to construct system,
where building blocks may be of any type.

@@

We will describe tree generating method described by Koza \cite{koza92}
called \textit{ramped half-and-half}. \\

@@

Generated individual is added into the initial population only if 
the initial population does not yet contain this individual.
This way it is guaranteed that every individual is unique 
in the initial population.

@@

In the field of evolutionary algorithms there is plenty of 
options for selection mechanisms 
for us to choose from. Again we will describe mechanism Koza
used in his first book on GP. It is the \textit{Roulette selection}.
It uses fitness value of each individual in straightforward way to determine probability of selecting this individual.\\
\newpage
Let there be $popSize$ individuals in the population.\\
And let $f_{i}$ be fitness value for individual i 
where $i \in \setDots{1}{popSize}$.

@@

In GP the most common crossover is \textit{Subtree swapping crossover}.
This crossover randomly selects one node in each parent tree.

With probability $p_{ip}$ (e.g. $p_{ip} = 90\%$) the crossover node
is randomly (uniformly) selected from internal nodes (i.e. function nodes) and
with probability $1 - p_{ip}$ (e.g 10\%) from leaf nodes.
This is done for both parents.

Two new child individuals are constructed by swapping subtrees 
that have roots in those selected nodes.\\

@@

And here are two child trees with swapped subtrees:\\

@@

@@

\section{Construction of next population}

@@

Evaluation is process of assigning fitness value to all 
individuals in the population. It is also usual to check
individuals for being correct solution.
If such a correct solution is find, the algorithm terminates. 
In our implementation fitness function returns
pair of real number (fitness value) and boolean value (indicating
whether tested individual is correct solution).

@@

Bellow is described GP algorithm in pseudocode. 
It is slight modification of the standard GP algorithm, 
because the best individual of the 
generation is preserved unchanged into the next generation.

\Pseudokod{GP( fitness, \TuF, popSize, numGens, probabs )}{
  $gen \leftarrow$ 0 \;
  $pop \leftarrow generateInitialPopuletion( $ \TuF$ ) $ \;
  ($popWithFitness,terminate,best) \leftarrow$ 
  evaluate($fitness$, $pop$)\;\;

@@

			\Case{Crossover}{
				$parent1 \leftarrow$ selection( $popWithFitness$ ) \;
				$parent2 \leftarrow$ selection( $popWithFitness$ ) \;\;

@@

				$newPop$.insert( $child1$ ) \;
				$newPop$.insert( $child2$ ) \;
				$i \leftarrow i + 2$ \;
			}
			\Case{Reproduction}{
				$indiv \leftarrow$ selection( $popWithFitness$ ) \;
				$newPop$.insert( $indiv$ ) \;
				$i \leftarrow i + 1$ \;
			}
			\Case{Mutation}{
				$indiv \leftarrow$ selection( $popWithFitness$ ) \;
				$mutant \leftarrow$ mutate( $indiv$ , \TuF ) \;
				$newPop$.insert( $mutant$ ) \;
				$i \leftarrow i + 1$ \;	
			}		
		}
	}\;

	$pop \leftarrow newPop$  \;
	($popWithFitness,terminate,best) \leftarrow$ 
	evaluate($fitness$, $pop$)\;
	$gen \leftarrow gen + 1$ \; 
 }\;

 \Return $popWithFitness$ \;
}

@@

Behaviors of contained procedures are hopefully clear from verbal description.
\\[2em]

For more detailed information on subject of GP and its many variants
see \cite{fg} containing many references to other works.

@@

\newcommand{\bbarr}{\twoheadrightarrow_\beta}
\newcommand{\barr}{\rightarrow_\beta}
\newcommand{\beq}{=_\beta}

\newcommand{\eearr}{\twoheadrightarrow_\eta}
\newcommand{\earr}{\rightarrow_\eta}
\newcommand{\eeq}{=_\eta}

\newcommand{\bearr}{\rightarrow_{\beta\eta}}
\newcommand{\bbeearr}{\twoheadrightarrow_{\beta\eta}}
\newcommand{\beeq}{=_{\beta\eta}}

@@

\begin{todo}
 & všude nahradit ty množinový za $\Gamma , x:\sigma$ a taky $(x,\Sigma)$ kde se to nehodí.
 & Až to bude dostatečně sepsáno, tak do této kapitoly taky přidat
   část o abstraction elimination a o polymorfním zobecnění inhabitation tree.
   Nezapomenout to pak zmínit i v tomhle úvodu.
\end{todo}

@@

We use bold upper case symbols such as 
$\Scomb, \Kcomb, \Icomb$ as abbreviation
for specific fixed terms, e.g. 
$\Icomb$ always means $\lamb{x}{x}$.  
\\[1em]

@@

$M[x:=N]$ the substitution of a term $N$ for the free occurrences of 
a variable $x$ in a term $M$ is defined inductively as follows.
\begin{align*}
x[x:=N]           &= N \\
y[x:=N]           &= y &\text{ where } x \not= y  \\
(P~Q)[x:=N]       &= ( P[x:=N]  ~ Q[x:=N] ) \\
\lamb{x}{P}[x:=N] &= \lam{x}{P}\\
\lamb{y}{P}[x:=N] &= \lam{y}{( P[x:=N] )} &\text{ where } x \not= y, y\not\in \FV(N)\\
\end{align*}

In order to satisfy the $y\not\in \FV(N)$ condition, names of bound variables
are chosen such that they differ from the free ones in the term.

@@

In other words, $\beta$-reduction is the process 
of insertion of arguments supplied to a function into 
its body. \\

Another useful relations are $\bbarr$ and $\beq$ defined as follows.

\begin{enumerate}
 \item \begin{enumerate}
 	\item $M \bbarr M$
 	\item $M \barr N \then M \bbarr N$
 	\item $M \bbarr N , N \bbarr L \then M \bbarr L$ 	
 \end{enumerate}
 \item \begin{enumerate}
 	\item $M \bbarr N \then M \beq N$
 	\item $M \beq N \then N \beq M$
 	\item $M \beq N , N \beq L \then M \beq L$
 \end{enumerate}

We read those relations as follows.
\begin{enumerate}
 	\item $M \bbarr N$ --- "$M$ $\beta$-reduces to $N$."  
 	\item $M \barr N$  --- "$M$ $\beta$-reduces to $N$
 	      in one step."
 	\item $M \beq N$ --- "$M$ is $\beta$-convertible to $N$."	
 \end{enumerate}

\begin{todo}
& definovat \textbf{CR} a říct o tom lemátku, že má CR redukce má unikátní nf.
&& V \cite{barendregt84} (ta nová kniha s žlutejma deskama) je ukázano že je eta CR (Church Rosserovská)
& zmínit Theorem 2B.4 z \cite{barendregt10}
\end{todo}

@@

$$\lamb{x}{(M~x)} \eArrow M \textbf{ ~~~~if } x \not\in FV(M) $$~

Analogically, a term of a form $\lamb{x}{(M~x)}$ is called 
\textit{$\eta$-redex}.

Relation $\bearr\;=\;\barr \cup \earr$. 
(Relation $R = \{\;(a,b)\;|\;a\;R\;b\;\}$.)

Similarly as for $\bbarr$ and $\beq$ we can define relations 
$\eearr$, $\eeq$,$\bbeearr$ and $\beeq$.

@@

$$M \eeArrow \lamb{x}{(M~x)} \textbf{ ~~~~if } x \not\in FV(M) $$

@@

~\begin{easylist}[enumerate]
& A \lterm is a \textit{$\beta$-normal form} (\bnf) 
if it does not have a $\beta$-redex as subterm.
& A \lterm M \textit{has} a \bnf if $M \beq N$
and $N$ is a \bnf.\\
\end{easylist}
A normal form may be thought of as a result of a term evaluation.

Similarly we can define \enf and \benf.

@@

\newcommand{\sexprTree}{sexpr-tree\xspace}
\newcommand{\SexprTree}{Sexpr-tree\xspace}

@@

@@

@@

In our system, the set of \textit{building blocks} 
(the equivalent of the set $T \cup F$ in standard GP)
will be a context with all declarations having \textit{constants}
as subjects, i.e., a set of constants together with their types.

In the process of term generation, 
this initial context is being further extend with variables
 --- by "entering inner scopes of bodies of lambda functions".\\

Notation $\Gamma,x:\sigma $ denotes $ \Gamma\cup\{(x:\sigma)\}$ 
such that $\Gamma$ does not contain any declaration with $x$ as subject.

\subsection{Simply typed $\lambda$-calculus}

@@

\begin{todo}
   & říct že se vždycky zastavěj (strong normalization)
     && a jaký to má výhody pro nás v GP (duležitý),
     && potřebný definice pro to uvést v \lterms sekci
     && musí se říct co je strongly normalizing, 
        tzn simpli typed $lambda$-kalkulus termy a $\beta$-redukce /
        $\beta\eta$-redukce
%     && řekl bych že eta redukce je SN vždycky (tzn i untyped lk)
%        pač plati (CR+terminující pak SN) a je CR vzhledem k barendrecht-stará-žlutá
%        a terminující je rpotože zmenšuje term s nemuže zmenšovat donekonečna    
\end{todo}

@@

\section{Term generating grammar}
\label{Term-generating-grammar}

@@

For every non-terminal $(\sigma \ar \tau,\Gamma)$
and for every variable $x$ such that $\Gamma$ does not contain any 
declaration with $x$ as subject
there is following grammar rule.

@@

$$ 
	( \sigma \rightarrow \tau , \Gamma )  \gar
	\bigg( \mbox{ {\Large $\lambda$ x . }}( \tau ; \Gamma ,x:\sigma ) \quad \bigg)
$$\\

@@

@@

@@

@@

@@

\Lets consider following definition of 
\textit{and-or tree}.
%\footnote{
%\textbf{TODO}: Mention that on [WIKI] there is more general definition, but 
%for our purposes is this one sufficient.}

@@

\newpage
\subsection{ Generating terms in \lnf by solving inhabitation tree }

@@

\begin{todo}
& skontrolovat ty dukazy pořádně sepsal sem 
  je dost rychle a mužou bejt divný. 
\end{todo}

\newpage
\subsection{In defense of inhabitation trees}
